Formalization of Typed Lua

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: some fixes in intro

\begin{abstract}
Dynamically typed languages such as Lua avoid static types in order
to have a simpler and more flexible language, because their programmers
to not need to bother with abstracting types that need to be validated
by a type checker. In contrast, statically typed languages provide
earlier detection of many defects, and a better framework for structuring
large programs. The advantages of static types often lead programmers to
migrate from a dynamically typed to a statically typed language when
their simple scripts evolve into complex programs.

Optional type systems are one way to have both static and dynamic
typing in the same language while keeping its original semantics,
making evolving a program from dynamic to static typing a matter of
describing the implied types that it it using and adding annotations
to make those types explicit. Designing an optional type system for
an existing dynamically typed language is challenging, as its types
should feel natural to programmers that are already familiar with
this language.

In this work, we give a formal description of Typed Lua, an
optional type system for Lua, with a focus on its novel type system
features: incremental evolution of record and object types via
imperative update that does not sacrifice width subtyping,
and combining flow typing with multiple assignment and with functions that
return multiple values. While our type system is tailored to the
features and idioms of Lua, its features can be adapted to other
imperative scripting languages.
\end{abstract}

@@

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing.
\emph{Optional type systems}~\cite{bracha2004pluggable} and
\emph{gradual typing}~\cite{siek2006gradual} are two similar
approaches for blending static and dynamic typing in the same
language. Their aim is to offer programmers the option
of adding type annotations where static typing is needed,
allowing the incremental migration from dynamic to static
typing. The difference between these two approaches is the
way they treat the runtime semantics of the language:
optional type systems preserve the original semantics
of a dynamically typed language, at the cost of runtime
safety, while gradual typing changes the semantics to
include more extensive runtime checking to ensure that
dynamically typed portions of the program do not violate
the type safety of statically typed portions.

Lua is a small imperative language with lexically-scoped first-class functions where the only data structuring
mechanism is the \emph{table} --
an associative array that has syntactic and runtime support
to efficiently represent arrays, records, maps, modules,
objects, etc. Unlike other scripting languages, Lua has very limited coercion among different data types.

Lua prefers to provide mechanisms instead of fixed policies due
to its primary use as an embedded language for configuration and
extension of other applications.
This means that even features such as a module system and
object orientation are a matter of convention instead of
built-in language constructs.
The result is a fragmented ecosystem of libraries, and different
ideas among Lua programmers on how they should use the language
features or on how they should structure programs.

Typed Lua is an optional type system for
Lua that is rich enough to preserve some of the idioms
that Lua programmers are already familiar with, while
helping programmers structure large Lua programs and
catch errors at compile-time. An informal
presentation of the design of Typed Lua has been
published~\cite{maidl2014tl}.

In this paper, we give a formal basis to an updated version of that design, with a focus on novel type system features
that Typed Lua uses to type some tricky Lua idioms.
One of these idioms is the use of assignment to
build records and objects one field at a time, as in
the following example, where the {\tt newpoint} function
is a constructor for objects with {\tt x} and {\tt y}
fields and a {\tt move} method:
\begin{verbatim}
  function new_point(x, y)
    local point = {}
    point.x = x
    point.y = y
    function point:move(dx, dy)
      self.x = self.x + dx
      self.y = self.y + dy
      return self
    end
    return point
  end
\end{verbatim}

The constructor function builds the object piece-by-piece,
from an initially empty table. Typed Lua
lets record and object types evolve incrementally 
via assignment while not sacrificing width subtyping.

Another idiom combines overloading on the return values
of a function, multiple return values, multiple assignment
to decompose multiple return values,
and tag checks to discriminate a several returned values
by checking just one of them, as in the following example:
\begin{verbatim}
  function idiv(x, y)
    if y == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end

  local q, r = idiv(a, b)
  if q then
    assert(a == q*b + r)
  else
    error(r)
  end
\end{verbatim}

@@

The lack of standard policies is a challenge for the design of
an optional type system for Lua. For this reason, the design
of Typed Lua has been informed by a mostly automated survey
of Lua idioms used by a large corpus of Lua libraries.
This paper also presents the methodology of this survey and
a summary of its results.

The rest of the paper is organized as follows:
Section \ref{sec:statistics} presents the methodology and results of
our survey of Lua idioms;
Section \ref{sec:types} presents the types of Typed Lua's
type system; Section~\ref{sec:subtyping} presents
their subtyping and consistent-subtyping relations;
Section \ref{sec:rules} presents the typing rules for evolution of table types,
and the typing rules for destructuring unions of tuples into projection types,
allowing flow typing on these projections;
Section \ref{sec:related} reviews related work;
Section \ref{sec:conclusion} presents our conclusions,
and outlines our plans for future work.

@@

Our first analysis surveyed table constructors (expressions
that create and initialize a table) to find out how a table
begins its life. Of all of the 23,185 table constructors of
our corpus, 36\% of them create a record (a table with string
keys such as \texttt{\{ x = 120, y = 121 \}}), 29\% of them create
a list or a tuple (a table with consecutive integer keys such as \texttt{\{ "one", "two", "three", "four" \}}), 26\% of them create an empty table, and 8\% of them create a table with
both a record part and a list part. The remaining 1\% of them create a table with non-literal keys.

@@

\begin{figure*}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure*}

@@

Table types represent some of the forms that Lua
tables can take. What a particular table type denotes
depends on its {\em tag}: the {\em unique}, {\em open}
and {\em fixed} tags all denote tables where it is
guaranteed that every key belongs to one of the {\em key
types} $K_i$. We restrict key types to literal types,
base types, and the top type to make the system simpler
for the user, as our survey of Lua programs showed that
this restriction reflects the way Lua tables are used.
Fields are assignable by default, but can be marked
as non-assignable with the {\bf const} qualifier. Having
a field be {\bf const} is not a guarantee that its
value cannot change, as the table may have aliases with
a non-{\bf const} type for that field.

The difference between {\em unique}, {\em open} and
{\em fixed} tables is the guarantees that the type system
give about the reference to a value of that type. A reference
to an {\em unique} table is guaranteed to have no aliases,
so allows the greater flexibility in reinterpreting its type;
a reference to an {\em open} table is guaranteed to have
only {\em closed} aliases (we will explain the the meaning of the {\em closed} tag shortly), which still lets the type system
reinterpret the type of a reference, but with more
restrictions; a {\em fixed} table can have any number of {\em fixed} or {\em closed} aliases, so its type cannot change
anymore.

@@

We present the subtyping rules as a deduction system for the
subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
The variable $\senv$ is a set of subtyping assumptions used
by the subtyping relation between recursive types.

Most of the subtyping rules are standard; we will focus
on the interesting cases, involving second-level types and
table types.

Tuple types are covariant, so the subtyping rule for pair types is straightforward:
\[
\begin{array}{c}
\mylabel{S-PAIR}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\]

>>>>

\begin{abstract}
Programmers often migrate from a dynamically typed
to a statically typed language when their simple scripts
evolve into complex programs. Optional type systems are
one way of having both static and dynamic typing
in the same language, while keeping its dynamically typed
semantics. This makes evolving a program from dynamic to
static typing a matter of describing the implied types that
it it using and adding annotations to make those types explicit.
Designing an optional type system for
an existing dynamically typed language is challenging, as its types should feel natural to programmers that are already familiar with
this language.

In this work, we give a formal description of Typed Lua, an
optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of record and object types via imperative update that does not sacrifice width subtyping, and combining flow typing with multiple assignment and with functions that return multiple values. While our type system is tailored to the features and idioms of Lua, its features can be adapted to other imperative scripting languages.
\end{abstract}

@@

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing.
\emph{Optional type systems}~\cite{bracha2004pluggable} and
\emph{gradual typing}~\cite{siek2006gradual} are two 
approaches for blending static and dynamic typing in the same
language. Their aim is to offer programmers the option
of adding type annotations where static typing is needed,
allowing the incremental migration from dynamic to static
typing. The difference between these two approaches is the
way they treat the runtime semantics of the language:
optional type systems preserve the original semantics
of a dynamically typed language, at the cost of runtime
safety, while gradual typing changes the semantics to
include more extensive runtime checking. These extra checks
ensure that
dynamically typed portions of the program do not violate
the type safety of statically typed portions.

Lua is a small imperative language with lexically-scoped first-class functions where the only data structuring
mechanism is the \emph{table}, an associative array. Lua prefers to provide mechanisms instead of fixed policies due
to its primary use as an embedded language for configuration and
extension of other applications.
This means that even features such as a module system and
object orientation are a matter of convention on how
programs use the syntactic sugar and metaprogramming
facilities of tables, instead of
built-in language constructs. Tables can efficiently represent arrays, records, maps, modules,
objects, etc.

The result is a fragmented ecosystem of libraries, and different
ideas among Lua programmers on how they should use the language
features or on how they should structure programs. One example
of this fragmentation is the proliferation of object systems.

Typed Lua~\cite{maidl2014tl} is an optional type system for
Lua that is rich enough to preserve some of the idioms
that Lua programmers are already familiar with, while
helping programmers structure large Lua programs and
catch errors at compile-time.

In this paper, we give a formal basis to an updated version of Typed Lua's original design, with a focus on two novel type system
features that Typed Lua uses to type some tricky Lua idioms.
The lack of standard policies is a challenge for the design of
an optional type system for Lua. For this reason, the design
of Typed Lua has been informed by a mostly automated survey
of Lua idioms used by a large corpus of Lua libraries.
This paper also presents the methodology of this survey and
a summary of its results.

The first idioms that we focus on this paper is the use of assignment to build records and objects one field at a time, as in
the Lua fragment in Figure~\ref{fig:newpoint}, where the {\tt newpoint} function
is a constructor for objects with {\tt x} and {\tt y}
fields and a {\tt move} method.

\begin{figure}[t]
\label{fig:newpoint}
\begin{verbatim}
  function new_point(x, y)
    local point = {}
    point.x = x
    point.y = y
    function point:move(dx, dy)
      self.x = self.x + dx
      self.y = self.y + dy
      return self
    end
    return point
  end
\end{verbatim}
\caption{Building an object piece-by-piece}
\end{figure}

The {\tt newpoint} function builds the object piece-by-piece,
from an initially empty table. Typed Lua
lets record and object types evolve incrementally 
via assignment while not sacrificing width subtyping.

The other idiom that we focus on combines overloading on the return values of a function, multiple return values, multiple assignment
to decompose multiple return values,
and tag checks to discriminate several returned values
by checking just one of them. The Lua fragment in Figure~\ref{fig:over} shows a function with overloaded multiple
returns, and how a caller of this function can discriminate
which list of multiple values the function returned.

\begin{figure}[t]
\label{fig:over}
\begin{verbatim}
  function idiv(x, y)
    if y == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end

  local q, r = idiv(a, b)
  if q then
    assert(a == q*b + r)
  else
    error(r)
  end
\end{verbatim}
\caption{Overloading on multiple returns}
\end{figure}

@@

The remainder of the paper is organized as follows:
Section \ref{sec:statistics} presents the methodology and results of
our survey of Lua idioms;
Section \ref{sec:types} presents the types of Typed Lua's
type system; Section~\ref{sec:subtyping} presents
their subtyping and consistent-subtyping relations;
Section \ref{sec:rules} presents the typing rules
for the two idioms that we focus on this paper;
Section \ref{sec:related} reviews related work;
Section \ref{sec:conclusion} presents our conclusions,
and outlines plans for future work.

@@

Our first analysis surveyed table constructors (expressions
that create and initialize a table) to find out how a table
begins its life. Of all of the 23,185 table constructors of
our corpus, 36\% of them create a record (a table with string
keys such as \texttt{\{x = 1, y = 2\}}), 29\% of them create
a list or a tuple (a table with consecutive integer keys such as \texttt{\{"one", "two", "three", "four"\}}), 26\% of them create an empty table, and 8\% of them create a table with
both a record part and a list part. The remaining 1\% of them create a table with non-literal keys.

@@

\begin{figure*}[!ht]
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; T \times P & \textit{pair types}
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

@@

Table types represent some of the forms that Lua
tables can take. What a particular table type denotes
depends on its {\em tag}: the {\em unique}, {\em open}
and {\em fixed} tags all denote tables where it is
guaranteed that every key belongs to one of the {\em key
types} $K_i$. We restrict key types to literal types,
base types, and the top type to make the system simpler
for the user, as our survey of Lua programs showed that
this restriction reflects the way Lua tables are commonly used.

Fields are assignable by default, but can be marked
as non-assignable with the {\bf const} qualifier. Having
a field be {\bf const} is not a guarantee that its
value cannot change, as the table may have aliases with
a non-{\bf const} type for that field. Non-assignable
fields enable {\em depth subtyping} between table types.

The difference between {\em unique}, {\em open} and
{\em fixed} tables is the guarantees that the type system
gives about the reference to a value of that type. A reference
to an {\em unique} table is guaranteed to have no aliases,
so it allows greater flexibility in reinterpreting its type;
a reference to an {\em open} table is guaranteed to have
only {\em closed} aliases (we will explain the meaning of the {\em closed} tag shortly), which still lets the type system
reinterpret the type of a reference, but with more
restrictions; a {\em fixed} table can have any number of {\em fixed} or {\em closed} aliases, so its type cannot change
anymore.

@@

The subtyping rules are a deduction system for the
subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
The variable $\senv$ is a set of subtyping assumptions used
by the subtyping relation between recursive types. 
Most of the subtyping rules are standard, so we will
focus on the interesting cases.

\begin{figure}
\label{fig:subtuples}
\[
\begin{array}{c}

\]
\caption{Subtyping for second-level types}	
\end{figure}

Figure~\ref{fig:subtuples} shows the subtyping rules for
second-level types. 
Tuple types are covariant, so the subtyping rule for pair types is straightforward:

%\[
%\begin{array}{c}
%\mylabel{S-PAIR}\\
%\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
%       \senv \vdash P_{1} \subtype P_{2}}
%      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
%\end{array}
%\]
